---
title: "Final"
author: "Ian Dalton"
date: "May 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(DT)
library(tidyr)
# library(corrplot)
# library(leaflet)
library(lubridate)
```

# Name TBD
I will think of a cool name later.  

walk users through the entire data science pipeline:
- data curation,
- parsing, and management
- exploratory data analysis
- hypothesis testing and
- machine learning to provide analysis
- and then the curation of a message or messages covering insights learned during the tutorial.


## Introduction  

**DATA**

Here I will discuss the data set I will be using for my final projects Analysis.

I will have embedded R chunks like this one which will show exactly what kind of data I will be using.

```{r cars}
summary(cars)
```

##Plots

there will also be plots because who doesnt like well formated plots?  

```{r pressure, echo=FALSE}
plot(pressure)
```

test 

```{r include=FALSE}
options(tibble.width = Inf)
```


Lets first have a look at these files:

### Peek at the dataset {.tabset}

#### Properties
There is a total of 58 features per property. You can scroll the x-axis to see all features.

```{r, result='asis', echo=FALSE}
datatable(cars, style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))
```

#### Transactions

```{r, result='asis', echo=FALSE}
datatable(pressure, style="bootstrap", class="table-condensed", options = list(dom = 'tp'))
```


Stay tuned!