---
title: "Final"
author: "Ian Dalton"
date: "May 18, 2018"
output: 
  html_document: 
    toc: yes
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(DT)
library(tidyr)
# library(corrplot)
# library(leaflet)
library(lubridate)
```

# Name TBD
I will think of a cool name later.  

walk users through the entire data science pipeline:  
- data curation,  
- parsing, and management  
- exploratory data analysis  
- hypothesis testing and  
- machine learning to provide analysis  
- and then the curation of a message or messages covering insights learned during the tutorial.  


## Introduction  

**DATA**

Here I will discuss the data set I will be using for my final projects Analysis.

I will have embedded R chunks like this one which will show exactly what kind of data I will be using.

```{r cars}
summary(cars)
```

##Plots

there will also be plots because who doesnt like well formated plots?  

```{r pressure, echo=FALSE}
plot(pressure)
```

test 

```{r include=FALSE}
options(tibble.width = Inf)
```


Lets test some demonstration of data

### example table {.tabset}

#### table1
There is an example of a dynamic data table possible because of web hosting the html output yippee.

```{r, result='asis', echo=FALSE}
datatable(cars, style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))
```

#### table2

```{r, result='asis', echo=FALSE}
datatable(pressure, style="bootstrap", class="table-condensed", options = list(dom = 'tp'))
```


Stay tuned!